---
title: "Fairness Analysis"
output: html_notebook
---

#libraries and load dataset

```{r}
library(tidyverse)
library(ggplot2)    
library(corrplot)    
library(caret)       
library(glmnet)     
library(pROC)
library(car)  
library(gtsummary)
library(scales)
library(patchwork)
library(RColorBrewer)
library(randomForest)
library(rpart)
library(rpart.plot)
library(rattle)
library(xgboost)
library(fairness)
```


```{r}
mydata <- read.csv("Pima Indians Diabetes.csv")
```


#dataset overview:


```{r}
summary(mydata)
```
```{r}
mydata %>%
  summarise(across(where(is.numeric), list(
    `Data Type` = ~class(.),
    `Distinct Values` = ~n_distinct(.),
    `Mean` = ~mean(., na.rm = TRUE),
    `Median` = ~median(., na.rm = TRUE),
    `Std Dev` = ~sd(., na.rm = TRUE),
    `Min` = ~min(., na.rm = TRUE),
    `Max` = ~max(., na.rm = TRUE),
    `Q1` = ~quantile(., 0.25, na.rm = TRUE),  # First Quartile (Q1)
    `Q3` = ~quantile(., 0.75, na.rm = TRUE)   # Third Quartile (Q3)
  ))) %>%
  pivot_longer(cols = everything(), names_to = c("Column", ".value"), names_sep = "_")



```



#Data Preparation: 

##change variable type


```{r}
mydata$diabetes <- factor(mydata$diabetes, levels = c(0, 1), labels = c("No", "Yes"))
```

```{r}
library(dplyr)

# Create a summary of diabetes type and its levels
diabetes_levels <- data.frame(
  Main_Level = c(1, 0),            # Main levels (1 and 0)
  Diabetes_level = c("Yes", "No"), # Diabetes level (Yes for 1, No for 0)
  Level = c("Diabetic", "Non-Diabetic")  # Descriptive level (Diabetic and Non-Diabetic)
)

# Print the table
print(diabetes_levels)





```


##NA values

```{r}
missing_values <- colSums(is.na(mydata))
print(missing_values)

```
```{r}
# Calculate the number of missing values in each column
missing_values <- colSums(is.na(mydata))

# Create a data frame for better visualization
missing_values_table <- data.frame(
  Column = names(missing_values),
  Missing_Values = missing_values
)

# Print the table
print(missing_values_table)

```

##Count zero values:

```{r}
zero_counts <- sapply(mydata[ , sapply(mydata, is.numeric)], function(x) sum(x == 0, na.rm = TRUE))

zero_summary <- data.frame(Variable = names(zero_counts), Zero_Count = zero_counts)

print(zero_summary)

```


##Remove zeros:

```{r}
invalid_zero_vars <- c("Glucose", "BloodPressure", "SkinThickness", "Insulin", "BMI")
mydata[invalid_zero_vars] <- lapply(mydata[invalid_zero_vars], function(x) {
  x[x == 0] <- NA
  return(x)
})

mydata[invalid_zero_vars] <- lapply(mydata[invalid_zero_vars], function(x) {
  x[is.na(x)] <- median(x, na.rm = TRUE)
  return(x)
})

colSums(is.na(mydata[invalid_zero_vars]))



```

##outliers:


```{r}
count_outliers <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE) 
  Q3 <- quantile(x, 0.75, na.rm = TRUE) 
  IQR_value <- Q3 - Q1                  
  lower_bound <- Q1 - 1.5 * IQR_value     
  upper_bound <- Q3 + 1.5 * IQR_value 
  
  sum(x < lower_bound | x > upper_bound, na.rm = TRUE)
}

outlier_summary <- mydata %>%
  summarise(across(where(is.numeric), count_outliers, .names = "Outliers_in_{.col}"))

print(outlier_summary)

```


# descriptive analytics


##pie chart for distribution of patients


```{r}
mydata %>%
  count(diabetes) %>%
  mutate(percentage = n / sum(n) * 100) %>%
  ggplot(aes(x = "", y = percentage, fill = diabetes)) +
  geom_bar(stat = "identity", width = 2, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = paste0(diabetes, "\n", round(percentage, 1), "%")),
            position = position_stack(vjust = 0.5), size = 7) +  
  scale_fill_manual(values = c("No" = "red", "Yes" = "green")) +
  labs(title = "Proportion of Diabetes Cases", x = "", y = "") +
  theme_void() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 14))

```


#Distribution of predictors:


```{r}

mydata_dist <- mydata %>%
  select(Pregnancies, Glucose, BloodPressure, SkinThickness, 
         Insulin, BMI, DiabetesPedigreeFunction, Age) %>%
  pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value")

ggplot(mydata_dist, aes(x = Value)) +
  geom_histogram(aes(y = ..density..), 
                 fill = "blue", color = "white", bins = 60, alpha = 0.5) +
  geom_density(color = "black", size = 1) +
  facet_wrap(~Variable, scales = "free", ncol = 4) +
  labs(title = "Distribution of Predictors", x = NULL, y = "Density") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))


```



#correlation matrix:


```{r}

numeric_data <- mydata %>%
  select(where(is.numeric))

cor_matrix <- cor(numeric_data, use = "complete.obs")

corrplot(cor_matrix,
         method = "color",        
         type = "lower",          
         addCoef.col = "black", 
         tl.col = "black",        
         tl.cex = 0.8,        
         number.cex = 0.65,     
         col = colorRampPalette(c("red", "white", "blue"))(200),  
         title = "Correlation Matrix of Predictors",
         mar =c(0, 0, 0, 0)) 


```



# Predictor's corelation to Age with trendlines:


```{r}
mydata_trend <- mydata %>%
pivot_longer(cols = c(Glucose, BloodPressure, Insulin, BMI),
               names_to = "Variable", values_to = "Value")

ggplot(mydata_trend, aes(x = Age, y = Value)) +
geom_point(alpha = 0.4) +
stat_smooth(method = "lm", aes(color = "linear"), se = FALSE) +
stat_smooth(method = "lm", formula = y ~ poly(x, 2), aes(color = "polynomial"), se = FALSE) +
stat_smooth(method = "nls",
              formula = y ~ a * log(x) + b,
              method.args = list(start = list(a = 1, b = 1)),
              aes(color = "logarithmic"), se = FALSE) +

  facet_wrap(~ Variable, scales = "free_y") +
  scale_colour_brewer(name = "Trendline", palette = "Set1") +
  labs(title = "Relationships Between Age and Key Health Predictors (Glucose, Blood Pressure, Insulin, BMI) with Trendlines"
,x = "Age", y = "Value") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

```



#Distribution of patients to have a overvview befor applying the bias evaluation


```{r}

mydata_long <- mydata %>%
  pivot_longer(cols = c(Age, Pregnancies, BMI, SkinThickness, BloodPressure, 
                        Insulin, DiabetesPedigreeFunction, Glucose), 
               names_to = "Variable", 
               values_to = "Value")

mydata_long %>%
  ggplot(aes(x = Value, fill = diabetes, color = diabetes)) +
  geom_density(alpha = 0.3) +
  facet_wrap(~Variable, scales = "free", ncol = 4) +   # Faceting the plots
  labs(title = "Distribution of Predictors by Diabetes Outcome", 
       x = NULL, y = "Density") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

```



#Logestic regresion:(linear model)

```{r}
set.seed(123)
trainIndex <- createDataPartition(mydata$diabetes, p = 0.8, list = FALSE)
train <- mydata[trainIndex, ]
test  <- mydata[-trainIndex, ]
```


```{r}
train_control <- trainControl(
  method = "repeatedcv",  
  number = 10,             
  repeats = 5,             
  classProbs = TRUE,       
  summaryFunction = twoClassSummary
)

```


```{r}
logistic_model <- train(
  diabetes ~ .,              
  data = train,              
  method = "glm",            
  family = "binomial",       
  trControl = train_control, 
  metric = "ROC"            
)
```



```{r}
logistic_model$results
```


```{r}
test$pred_lr <- predict(logistic_model, test)                        
test$prob_lr <- predict(logistic_model, test, type = "prob")

```


```{r}
confusionMatrix_lr <-confusionMatrix(test$pred_lr, test$diabetes, positive = "Yes")
confusionMatrix_lr
```



```{r}
roc_curve_logestic <- roc(test$diabetes, test$prob_lr$Yes)
plot(roc_curve_logestic, col = "blue", main = "ROC_AUC Curve (Logestic Regression)", print.auc = TRUE)

```


#fairness analyses in logestic regression model:



```{r message=FALSE, warning=FALSE}
roc_auc_par_lr <- roc_parity(data         = test, 
                      outcome      = "diabetes", 
                      group        = "Age",
                      probs        = test$prob_lr$Yes, 
                     group_breaks = c(20, 35, 55, 80))


roc_auc_par_lr$Metric

```

```{r}
roc_auc_par_lr$ROCAUC_plot + 
  ggtitle("ROC AUC Comparison by Group (Logistic Regression Model)") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

```



##Predictive rate parity:

```{r}
pred_par_lr <- pred_rate_parity(
  data = test,                     
  outcome = "diabetes",           
  outcome_base = "Yes",           
  group = "Age",                  
  probs = test$prob_lr$Yes,  
  cutoff = 0.5,                    
   group_breaks = c(20, 40, 60, 81) 
)

pred_par_lr$Metric

```


```{r}
pred_par_lr$Metric_plot

```

##Demographic parity:

```{r}
dem_par_lr <- dem_parity(data         = test, 
                      outcome      = "diabetes",
                      outcome_base = "Yes", 
                      group        = "Age",                  
                      probs        = test$prob_lr$No,      
                      cutoff       = 0.5,                   
                      group_breaks = c(20, 40, 60, 81))    

dem_par_lr$Metric
```

```{r}
levels(pred_par_lr$Group)

```


##Accuracy parity


```{r}
acc_par_lr <- acc_parity(data    = test, 
                      outcome = "diabetes", 
                      group   = "Age",
                      probs   = test$prob_lr$Yes, 
                      cutoff  = 0.5, 
                       group_breaks = c(20, 40, 60, 81))
acc_par_lr$Metric
```

##Specificity parity:

```{r}
sp_par_lr <- spec_parity(data         = test, 
                      outcome      = "diabetes", 
                      outcome_base = 'Yes', 
                      group        = "Age",
                      probs        = test$prob_lr$Yes, 
                      cutoff       = 0.5, 
                      group_breaks = c(20, 40, 60, 81))
sp_par_lr$Metric
```


#Random forest:


```{r}
rf_model <- train(
  diabetes ~ ., 
  data = train, 
  method = "rf",
  metric = "ROC",
  trControl = train_control,
  tuneLength = 7,
  ntree=400
  
)

print(rf_model)
```


```{r}
rf_model$results
```


```{r}
test$pred_rf <- predict(rf_model, test)                        
test$prob_rf <- predict(rf_model, test, type = "prob")

```



```{r}
confusionMatrix_rf <-confusionMatrix(test$pred_rf, test$diabetes, positive = "Yes")
confusionMatrix_rf
```



```{r}
roc_curve_rf <- roc(test$diabetes, test$prob_rf$Yes)
plot(roc_curve_rf, col = "blue", main = "ROC_AUC Curve(Random Forest)", print.auc = TRUE)

```

#fairness analyses in random forest model:

```{r warning=FALSE}
roc_auc_par_rf <- roc_parity(data         = test, 
                      outcome      = "diabetes", 
                      group        = "Age",
                      probs        = test$prob_rf$Yes, 
                     group_breaks = c(20, 35, 55, 81))
roc_auc_par_rf$Metric
```



```{r}
roc_auc_par_rf$ROCAUC_plot + 
  ggtitle("ROC AUC Comparison by Group (Random Forest Model)") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))

```



##Predictive rate parity

```{r}
pred_par_rf <- pred_rate_parity(
  data = test,                     
  outcome = "diabetes",           
  outcome_base = "Yes",           
  group = "Age",                  
  probs = test$prob_rf$Yes,  
  cutoff = 0.5,                    
   group_breaks = c(20, 40, 60, 81) 
)
pred_par_rf$Metric

```

```{r}
pred_par_rf$Metric_plot

```


##Demographic parity


```{r}
dem_par_rf <- dem_parity(data      = test, 
                      outcome      = "diabetes",
                      outcome_base = "Yes", 
                      group        = "Age",                  
                      probs        = test$prob_rf$Yes,      
                      cutoff       = 0.5,                   
                      group_breaks = c(20, 40, 60, 81))    


dem_par_rf$Metric
```


##Accuracy parity


```{r}
acc_par_rf <- acc_parity(data    = test, 
                      outcome = "diabetes", 
                      group   = "Age",
                      probs   = test$prob_rf$Yes, 
                      cutoff  = 0.5, 
                       group_breaks = c(20, 40, 60, 81))
acc_par_rf$Metric
```


##Specificity parity

```{r}
sp_par_rf <- spec_parity(data         = test, 
                      outcome      = "diabetes", 
                      outcome_base = 'Yes', 
                      group        = "Age",
                      probs        = test$prob_rf$Yes, 
                      cutoff       = 0.5, 
                      group_breaks = c(20, 40, 60, 81))
sp_par_rf$Metric
```



#XGB:


```{r}
XGB_model <- train(
  diabetes ~ .,            
  data = train,            
  method = "xgbTree",     
  trControl = train_control,
  tuneLength = 3,          
  metric = "ROC"
)

print(XGB_model)
print(XGB_model$bestTune)
```


```{r}

test$xgb_pred <- predict(XGB_model, test)
test$xgb_prob <- predict(XGB_model, test, type = "prob")
confusionMatrix(test$xgb_pred, test$diabetes, positive = "Yes")
```



```{r}
roc_curve_xgb <- roc(test$diabetes,test$xgb_prob$Yes)
plot(roc_curve_xgb, col = "blue", main = "ROC_AUC Curve(XGB Model)", print.auc = TRUE)

```
##Fairness analysis for XGB:


```{r warning=FALSE}
roc_auc_par_xgb <- roc_parity(data         = test, 
                      outcome      = "diabetes", 
                      group        = "Age",
                      probs        = test$xgb_prob$Yes, 
                     group_breaks = c(20, 35, 55, 81))
roc_auc_par_xgb$Metric
```
```{r}
roc_auc_par_xgb$ROCAUC_plot + 
  ggtitle("ROC AUC Comparison by Group (XGBoost Model)") +
  theme(plot.title = element_text(size = 14, face = "bold", hjust = 0.5))
```


##Predictive rate parity

```{r}
pred_par_xgb <- pred_rate_parity(
  data = test,                     
  outcome = "diabetes",           
  outcome_base = "Yes",           
  group = "Age",                  
  probs = test$xgb_prob$Yes,  
  cutoff = 0.5,                    
   group_breaks = c(20, 40, 60, 81) 
)
pred_par_xgb$Metric

```

```{r}
pred_par_xgb$Metric_plot

```


##Demographic parity


```{r}
dem_par_xgb <- dem_parity(data      = test, 
                      outcome      = "diabetes",
                      outcome_base = "Yes", 
                      group        = "Age",                  
                      probs        = test$xgb_prob$Yes,      
                      cutoff       = 0.5,                   
                      group_breaks = c(20, 40, 60, 81))    


dem_par_xgb$Metric
```


##Accuracy parity


```{r}
acc_par_xgb <- acc_parity(data    = test, 
                      outcome = "diabetes", 
                      group   = "Age",
                      probs   = test$xgb_prob$Yes, 
                      cutoff  = 0.5, 
                       group_breaks = c(20, 40, 60, 81))
acc_par_xgb$Metric
```


##Specificity parity

```{r}
sp_par_xgb <- spec_parity(data         = test, 
                      outcome      = "diabetes", 
                      outcome_base = 'Yes', 
                      group        = "Age",
                      probs        = test$xgb_prob$Yes, 
                      cutoff       = 0.5, 
                      group_breaks = c(20, 40, 60, 81))
sp_par_xgb$Metric
```


# feature importance: 


```{r}
feature_importance_lr <- varImp(logistic_model)
print(feature_importance_lr)
```



```{r}
feature_importance_rf <- varImp(rf_model)
print(feature_importance_rf)
```


```{r}
feature_importance_xgb <- varImp(XGB_model)
print(feature_importance_xgb)
```



```{r}

logistic_importance_sorted <- feature_importance_lr$importance[,1][order(feature_importance_lr$importance[,1], decreasing = FALSE)]
logistic_features_sorted <- rownames(feature_importance_lr$importance)[order(feature_importance_lr$importance[,1], decreasing = FALSE)]

barplot(logistic_importance_sorted, 
        names.arg = logistic_features_sorted, 
        main = "Feature Importance - Logistic Regression", 
        col = "skyblue", 
        horiz = TRUE, 
        las = 1, 
        cex.names = 0.5)

rf_importance_sorted <- feature_importance_rf$importance[,1][order(feature_importance_rf$importance[,1], decreasing = FALSE)]
rf_features_sorted <- rownames(feature_importance_rf$importance)[order(feature_importance_rf$importance[,1], decreasing = FALSE)]

barplot(rf_importance_sorted, 
        names.arg = rf_features_sorted, 
        main = "Feature Importance - Random Forest", 
        col = "lightgreen", 
        horiz = TRUE, 
        las = 1, 
        cex.names = 0.5)

xgb_importance_sorted <- feature_importance_xgb$importance[,1][order(feature_importance_xgb$importance[,1], decreasing = FALSE)]
xgb_features_sorted <- rownames(feature_importance_xgb$importance)[order(feature_importance_xgb$importance[,1], decreasing = FALSE)]

barplot(xgb_importance_sorted, 
        names.arg = xgb_features_sorted, 
        main = "Feature Importance - XGB", 
        col = "lightcoral", 
        horiz = TRUE, 
        las = 1, 
        cex.names = 0.5)

```


